% Chapter Template

\chapter{Evaluation Plan}\label{ch:evaluation}

\section{Machine Readable Licence Representations}

% we are looking to eavaluate the project - a lot of this should go into the intro probably...

% measures will be cost (smart contracts!), lawyer avoidance, and advice from people that know about this stuff

Other goals of this project require that a legal agreement can be encoded for it to become structured data.
The project should come up with, or find in the literature and adapt, a suitable format for a legal agreement.
In order to limit the scope of the project, we will narrow the agreement to be encoded to licences.

A successful project should include:
\begin{itemize}
    \item \textbf{Machine-readable representation of licenses}.
    \item \textbf{Prototype Software} that uses the contract representation to ease the existing manual workflow related
    to agreements in a business.
    This should include a searchable database that is aware of licenses related to the same data or software.
    The software should, to some extent, allow querying a contract to be able to determine the conditions which the
    license allows to use the data in.
    \item Secure, self-executing clauses that all parties bound to the contract can trust.
    This includes making it so parties cannot have plausible deniability or repudiation they are dishonest.
\end{itemize}

Legal contracts are complex documents that may list vague or hard-to-encode conditions and situations.
I expect not being able to (inexpensively) capture the entire meaning of a contract in a machine-readable format.
This disassociation between reality and representation should be fully taken into account when providing guarantees
and making assumption.
For example, a successful prototype should be aware enough to refer the user to the original license when it is unable
to provide certainties with respect to a query against a license.

In other words, this uncertainty created by attempting to represent a contract in a machine-readable encoding
should be part of the encoding itself, for the sake of completeness and usability.



