% Chapter Template

\chapter{Project Plan}\label{ch:timeline}



\section{Core Goals for Prototype}\label{sec:core-goals}

At the time of writing the interim report, the goals of the prototype are vague: I have not estimated
how long each idea will take to implement nor separated them into core goals and extensions.

In particular, it is yet to be decided how a licence will be encoded in a machine-readable format.
If this project decides to  tackle that, it wil be one of its main contributions.
If on the contrary it finds a suitable representation in the literature (like the Accord
project, see~\ref{subsec:accord}) then it can focus more other areas, like private verification through
zero-knowledge proofs or decentralised auditing through blockchains.


\section{Extensions}\label{sec:extensions}

Suitable extensions could include:
\begin{itemize}
    \item Going beyond encoding a single contract and reasoning about a party bound to several
    licences (some possibly related to the same data) and how these interact with each other.
    \item Improving ease-of-use of dataset-related licences by somehow relating data to the licence
    itself, in order to make the license easy to find.
\end{itemize}
% see https://tex.stackexchange.com/questions/196794/how-can-you-create-a-vertical-timeline
\newcommand{\foo}{\color{darkgray}\makebox[0pt]{\small\textbullet}\hskip-0.5pt\vrule width 1pt\hspace{\labelsep}}
\newcommand{\bfoo}{\raisebox{2.1ex}[0pt]{\makebox[\dimexpr2\tabcolsep]
{\color{LightSteelBlue3}\tiny\textbullet}}}
\newcommand{\tfoo}{\makebox[\dimexpr2\tabcolsep]{}}%
\daymonth

\newcommand{\para}{\newline\newline}

\renewcommand\arraystretch{1.5}
\captionsetup{font=blue, labelfont=sc, labelsep=quad}
\begin{longtable}{@{\,}r <{\hskip 2pt} !{\foo} >{\raggedright\arraybackslash}p{7cm}}
    \caption{Timeline} \\[-1.5ex]
    \toprule
    \addlinespace[1.5ex]
%    \multicolumn{1}{c!{\bfoo}}{}          & \\ [-3.4ex]
    \monthdate{27}{01}                    & \textbf{Interim report due}              \\
    \monthdate{27}{01}-\monthdate{10}{01} & Interim report review                    \\

    \monthdate{01}{03} & Acceptance criteria for software prototype defined.\para
    I should have decided the scope of the software prototype and settled for what
    representation this project will use to encode a contract. \newline
    I should have also started seeking out stakeholders willing to provide feedback during
    development - which will also be critical during the evaluation phase \\

    \monthdate{02}{03}-\monthdate{15}{04} & Bulk of prototype software development   \\

    \monthdate{16}{04} & Revision of software acceptance criteria \para
    I should refrain from large changes to the prototype beyond this point.
    I should revisit the project's goals for the prototype and decide if I have time for an extension
    or if on the contrary I need to further narrow the scope. \\
    \monthdate{17}{04}-\monthdate{09}{05} & Extension developing and                 \\

    \monthdate{10}{05}                    & \textbf{Project health checkup}          \\
    \monthdate{01}{05}-\monthdate{15}{05} & Project Evaluation \para
    I should have a prototype good enough to be evaluated by stakeholders in industry by May.
    I should implement critical feedback without fundamental code changes during this period. \\
    \monthdate{16}{05}-\monthdate{30}{05} & Report writing                           \\
    \monthdate{30}{05}                    & Report draft first hand-in               \\
    \monthdate{01}{06}-\monthdate{19}{06} & Address report feedback                  \\
    \monthdate{20}{06}                    & \textbf{Final report due}                \\
    \monthdate{24}{06}                    & \textbf{Slides and software archive due} \\
%    \multicolumn{1}{c!{\bfoo}}{} &
\end{longtable}

